---
title: "Reading, Checking, and Searching Text Data in R"
subtitle: "A readr and stringr lesson using NYC 311 complaints"
format:
  html:
    toc: true
    code-fold: true
    theme: cosmo
execute:
  warning: false
  message: false
---

## Lesson overview

This beginner-friendly lesson uses four tidyverse packages:

- **`readr`** imports, inspects, validates, and exports data.
- **`dplyr`** transforms, filters, and summarises data frames.
- **`tidyr`** reshapes data and handles missing values.
- **`stringr`** searches, measures, and transforms text.

We will use a public sample from NYC Open Data's 311 Service Requests. It has
text fields such as complaint type, descriptor, location type, and resolution
description, as well as missing values. 

### Learning objectives

Students will be able to:

1. Import a CSV from a URL with `readr::read_csv()`.
2. Inspect dimensions, names, types, and parsing problems.
3. Identify missing values and explain why they matter.
4. Use common `stringr` functions.
5. Create a new field with case-insensitive string matching.
6. Check whether a text-matching rule makes sense.
7. Export revised data with `readr::write_csv()`.

## 1. Packages and public data

```{r}
# Install once, if necessary:
# install.packages("tidyverse")

library(tidyverse)
```

This URL requests 15,000 rows and twelve useful fields. Because the public dataset
is updated over time, the exact rows may change when the lesson is rerun.

```{r}
nyc311_url <- paste0(
  "https://data.cityofnewyork.us/resource/erm2-nwe9.csv?",
  "$limit=15000&",
  "$select=created_date,closed_date,complaint_type,descriptor,",
  "location_type,incident_zip,borough,status,resolution_description,",
  "agency,agency_name,police_precinct"
)


nyc311_url
```

## 2. Import and check the CSV

```{r}
complaints_raw <- read_csv(
  file = nyc311_url,
  na = c("", "NA", "N/A")
)

complaints_raw
```

**The first row is used as the header by default**, and the result is a tibble.

```{r}
glimpse(complaints_raw)
```


Inspect types and parsing problems:

- `spec(complaints_raw)` shows the full column specification that `readr` used when importing — the name and **guessed** type of every column.


```{r}
spec(complaints_raw)
```


## 4. Common `readr` functions

| Function | Purpose |
|---|---|
| `read_csv()` | Read comma-separated files |
| `read_tsv()` | Read tab-separated files |
| `read_delim()` | Read a file with a chosen delimiter |
| `write_csv()` | Write a CSV file |
| `problems()` | Show parsing problems |
| `spec()` | Show column specifications |
| `parse_number()` | Extract numbers from text |
| `parse_date()` | Convert text into dates |

```{r}
parse_number(c("$1,250", "$950", "unknown"))
```

## Stringr lesson: searching and transforming text

## 5. Common `stringr` functions

```{r}
example_text <- c(
  "Loud Music/Party",
  "Noise - Residential",
  "Blocked driveway",
  NA
)

example_text
```

```{r}
str_length(example_text)                         # character count
str_to_lower(example_text)                      # lowercase
str_to_upper(example_text)                      # uppercase
str_detect(example_text, regex("noise", ignore_case = TRUE))
str_replace_all(
  example_text,
  regex("noise", ignore_case = TRUE),
  "SOUND"
)
str_extract(example_text, regex("noise|blocked", ignore_case = TRUE))
str_count(example_text, regex("noise", ignore_case = TRUE))
str_split("noise, parking, sanitation", pattern = ", ")
```

`str_detect()` returns `TRUE`, `FALSE`, or sometimes
`NA`. `regex()` allows case-insensitive matching. The `|`
symbol means “OR.”

## 6. Create a searchable text field

Combine several text columns, replacing missing values with empty strings first:

```{r}
complaints <- complaints_raw |>
  mutate(
    complaint_type_text = replace_na(complaint_type, ""),
    descriptor_text     = replace_na(descriptor, ""),
    location_type_text  = replace_na(location_type, ""),
    search_text = str_c(
      complaint_type_text,
      descriptor_text,
      location_type_text,
      sep = " | "
    )
  )

complaints |>
  select(search_text) |>
  slice_head(n = 6)
```

The `search_text` column is derived data created for a specific question.

## 7. Create a field using string matching

Identify complaints involving noise, loud sounds, music, or parties:

```{r}
complaints <- complaints |>
  mutate(
    is_noise_related = str_detect(
      search_text,
      regex("noise|loud|music|party|parties", ignore_case = TRUE)
    )
  )

complaints |> count(is_noise_related)
```

Inspect both matching and nonmatching records:

```{r}
complaints |>
  filter(is_noise_related) |>
  select(complaint_type, descriptor, location_type, is_noise_related) |>
  slice_head(n = 10)

complaints |>
  filter(!is_noise_related) |>
  select(complaint_type, descriptor, location_type, is_noise_related) |>
  slice_head(n = 10)
```

A matching rule can be syntactically correct but classify records incorrectly.
Always inspect examples.

## 8. Create broader categories

Conditions are checked from top to bottom, so order matters.

```{r}
complaints <- complaints |>
  mutate(
    topic = case_when(
      str_detect(search_text, regex("noise|loud|music|party|parties", ignore_case = TRUE)) ~ "noise",
      str_detect(search_text, regex("parking|vehicle|blocked driveway", ignore_case = TRUE)) ~ "parking",
      str_detect(search_text, regex("heat|hot water|water leak|plumbing", ignore_case = TRUE)) ~ "housing or water",
      .default = "other"
    )
  )

complaints |> count(topic)
```

These are not official NYC classifications. They are rule-based fields designed
for a particular analysis.

## 9. Text summaries and export

```{r}
complaints <- complaints |>
  mutate(
    descriptor_length  = str_length(descriptor_text),
    noise_word_count   = str_count(search_text, regex("noise", ignore_case = TRUE))
  )

complaints |>
  summarise(
    min    = min(descriptor_length),
    median = median(descriptor_length),
    mean   = round(mean(descriptor_length), 1),
    max    = max(descriptor_length)
  )

complaints |> count(noise_word_count)
```

```{r}
write_csv(complaints, "nyc311_text_classified.csv")
file.exists("nyc311_text_classified.csv")
```

Keep the original URL, download date, and matching rules with the project.

## Student practice activities

### Activity 1: Import and validate

1. Import the CSV into `complaints`.
2. Display rows, columns, names, and types.
3. Use `problems()` to check parsing.
4. Count missing values in every column.

### Activity 2: Explore text

Using `descriptor`:

1. Calculate the length of each descriptor.
2. Find the ten longest descriptors.
3. Count descriptors containing `"noise"`.
4. Repeat without regard to capitalization.
5. Explain how missing descriptors affect `str_detect()`.

### Activity 3: Create a new field

Create `is_parking_related` that is `TRUE` when the combined
text contains `"parking"`, `"vehicle"`, `"blocked driveway"`,
or `"illegal parking"`. Inspect matching and nonmatching rows and
identify one possible false positive or false negative.

### Activity 4: Improve the rule

Create `service_area` with these categories:

- `"noise"` for noise, loud, music, or party;
- `"parking"` for parking, vehicle, or driveway;
- `"housing"` for heat, water, plumbing, or apartment;
- `"other"` otherwise.

Explain why the order of the matching rules matters.

## Exit ticket

1. What is the difference between a missing value and a parsing problem?
2. Why convert text to lowercase before using `str_detect()`?
3. What does `|` mean inside `"noise|music|party"`?
4. Why inspect examples after creating a text-based classification?
5. What should be preserved so another person can reproduce the new field?

## Key takeaways

```{r}
#| eval: false
library(tidyverse)

data <- read_csv("URL_OR_FILE_PATH", show_col_types = FALSE)

glimpse(data)
spec(data)
problems(data)

data |> summarise(across(everything(), \(x) sum(is.na(x)))) |>
  pivot_longer(everything(), names_to = "column", values_to = "n_missing")

data <- data |>
  mutate(
    text_clean = replace_na(text_column, ""),
    new_field  = str_detect(text_clean, regex("pattern1|pattern2", ignore_case = TRUE))
  )
```

Importing data is not the same as validating data. Text matching is not the same
as discovering an objective truth: it is a rule that we create, test, explain,
and revise.

